Simple repression draft

### 2020-05-21 doc/section_01_introduction.tex
MRM 20200520 pass through abstract and intro

Gene expression presides over much of the most important dynamism 
of living organisms.   The level of expression of batteries of
different genes is altered as a result of spatiotemporal cues that integrate
chemical, mechanical and other types of signals.  The original repressor-operator model
conceived by Jacob and Monod in the context of bacterial
metabolism has now been transformed into the much
broader subject of gene regulatory networks in living organisms
of all kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining outstanding challenges
to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of
different regulatory architectures, stemming first and foremost
from our ignorance about what those architectures even are with
more than 60\% of the genes even in an ostensibly well understood organism
such as  {\it E. coli} with no regulatory
insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019, Santos_Zavaleta2019}.
But even once we have established the identity of key transcription
factors and their binding sites of a given promoter architecture,  there remains the predictive challenge
of understanding its input-output properties, an objective that can be met
by a myriad of approaches using the tools of statistical physics~\cite{Ackers1982, Shea1985, Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012, Saiz2013,
Ko1991,Peccoud1995,Record1996,
Kepler2001, Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root
to such predictive understanding is to focus on the simplest
regulatory architecture and to push the theory-experiment
dialogue as far and as hard as it can be pushed~\cite{Phillips2019}.  If we
demonstrate that we can pass that test by successfully predicting
both the means and variance in gene expression, then that provides
a more solid foundation upon which to launch into 
more complex problems.

To that end, in this paper we examine a wide variety of distinct models
for the simple repression regulatory architecture.  All of these
models coarse-grain away some of the important microscopic
features of these architectures that have been elucidated
by generations of genetics, molecular biology and biochemistry.  
Our aim is to understand the underlying principles of such
coarse-graining such that future models of regulatory response
can serve the powerful predictive role needed to take synthetic
biology from what is often a brilliant exercise in enlightened 
empiricism and recast it as the same kind of rational design process
we are used to in other branches of engineering.  
More precisely, we want phenomenology in the sense of
coarse-graining away atomistic detail, but still retaining biophysical meaning.
For example, we are not satisfied with the strictly phenomenological
approach offered by 
 Hill functions since we are motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
clearly insufficient since each new situation
requires a completely new set of parameters. Such work requires a quantitative theory of how biophysical
changes at the molecular level propagate to input-output functions at the
genetic circuit level. We want concepts, not mere facts. In particular a key
question is: at this level of coarse graining, what microscopic details do we
need to explicitly model, and how do we figure that out? For example, do we need
to worry about all or even any of the steps that individual RNAPs go through
each time they make a transcript? Turning the question around, can we see any
imprint of those processes in the available data? If the answer is no, then
those processes are irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling can complement each other beautifully here.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{\textbf{An overview of the simple repression motif at the level of
means.} (A) Schematic of the qualitative biological picture of simple repression. (B)
A variety of possible mathematicized cartoons of simple repression, along
with the effective parameter $\rho$ which subsumes all regulatory details of the
architecture that do not directly involve the repressor. (C) The ``master curve'' to which
all cartoons in (B) collapse.}
\label{fig1:means_cartoons}
\end{figure}

The logic of the remainder of the paper is as follows.
In section~\ref{section_02_means.tex}, we show how
both thermodynamic models and kinetic models based upon
the chemical master equation all culminate in the same underlying
functional form for the fold-change in the average level of gene
expression as shown in Figure~\ref{fig1:means_cartoons}(C).  Section~\ref{section_03_beyond_means.tex} goes
beyond an analysis of the means by asking how the same models
presented in Figure~\ref{fig1:means_cartoons} can be used
to explore noise in gene expression. Of course, to make contact
with experiment, all of these models must make a commitment
to some numerical values for the key parameters found in each
such model and in Section~\ref{section_04_bayesian_inference.tex}
we explore the use of Bayesian inference to establish
these parameters and to rigorously answer the question of how
to discriminate between the different models.

%
%\mmnote{Key ideas, no particular order, that I haven't written down before:
%\begin{itemize}
%\item Our goal is to build phenomenological models of input-output functions of
%genetic circuits. More precisely, we want phenomenology in the sense of
%coarse-graining away atomistic detail, but still retaining biophysical meaning,
%e.g., we don't want to coarse-grain as far as Hill functions. Why not? We are
%motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
%insufficient. Such work requires a quantitative theory of how biophysical
%changes at the molecular level propagate to input-output functions at the
%genetic circuit level. We want concepts, not mere facts. In particular a key
%question is: at this level of coarse graining, what microscopic details do I
%need to explicitly model, and how do we figure that out? For example, do I need
%to worry about all or even any of the steps that individual RNAPs go through
%each time they make a transcript? Turning the question around, can we see any
%imprint of those processes in the available data? If the answer is no, then
%those processes are irrelevant for our purposes. Forward modeling and inverse
%(statistical inferential) modeling can complement each other beautifully here.
%\end{itemize}}

>>>>

Gene expression presides over much of the most important dynamism of living
organisms.   The level of expression of batteries of different genes is altered
as a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals.  The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures. This challenge stems first and foremost from our ignorance about
what those architectures even are with more than 60\% of the genes even in an
ostensibly well understood organism such as  {\it E. coli} with no regulatory
insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
Santos_Zavaleta2019}. But even once we have established the identity of key
transcription factors and their binding sites of a given promoter architecture,
there remains the predictive challenge of understanding its input-output
properties, an objective that can be met by a myriad of approaches using the
tools of statistical physics~\cite{Ackers1982, Shea1985,
Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012,
Saiz2013, Ko1991,Peccoud1995,Record1996, Kepler2001,
Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
successfully predicting both the means and variance in gene expression, then
that provides a more solid foundation upon which to launch into more complex
problems.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
away some of the important microscopic features of this architecture that have
been elucidated by generations of geneticists, molecular biologists and
biochemists.
\mrm{Our aim is to understand the underlying principles of such coarse-graining
in order to increase the predictive power of these models. In doing so we hope
future models of regulatory response will be able to serve the powerful
predictive role needed to take synthetic biology from a brilliant exercise in
enlightened empiricism to a rational design framework as any other branch of
engineering.}
% Our aim is to understand the underlying principles of such
% coarse-graining such that future models of regulatory response can serve the
% powerful predictive role needed to take synthetic biology from what is often a
% brilliant exercise in enlightened empiricism and recast it as the same kind of
% rational design process we are used to in other branches of engineering. 
More precisely, we want phenomenology in the sense of coarse-graining away
atomistic detail, but still retaining biophysical meaning. For example, we are
not satisfied with the strictly phenomenological approach offered by the
commonly used Hill functions. Studies like~\cite{Razo-Mejia2018} have
demonstrated that Hill functions are clearly insufficient since each new
situation requires a completely new set of parameters. Such work requires a
quantitative theory of how biophysical changes at the molecular level propagate
to input-output functions at the genetic circuit level. We want concepts, not
mere facts. In particular a key question is: at this level of coarse graining,
what microscopic details do we need to explicitly model, and how do we figure
that out? For example, do we need to worry about all or even any of the steps
that individual RNAPs go through each time they make a transcript? Turning the
question around, can we see any imprint of those processes in the available
data? If the answer is no, then those processes are irrelevant for our purposes.
Forward modeling and inverse (statistical inferential) modeling are necessary to 
tackles such questions.

@@

\afterpage{\clearpage}
\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig01.pdf}
\caption{\textbf{An overview of the simple repression motif at the level of
means.} (A) Schematic of the qualitative biological picture of the simple
repression genetic architecture. (B) and (C) A variety of possible
mathematicized cartoons of simple repression, along with the effective parameter
$\rho$ which subsumes all regulatory details of the architecture that do not
directly involve the repressor. (B) Simple repression models from an equilibrium
perspective. (C) Equivalent models cast in chemical kinetics language. (D) The
``master curve'' to which all cartoons in (B) and (C) collapse.}
\label{fig1:means_cartoons}
\end{figure}

The logic of the remainder of the paper is as follows. In
section~\ref{section_02_means}, we show how both thermodynamic models and
kinetic models based upon the chemical master equation all culminate in the same
underlying functional form for the fold-change in the average level of gene
expression as shown in Figure~\ref{fig1:means_cartoons}(D).
Section~\ref{section_03_beyond_means.tex} goes beyond an analysis of the mean
gene expression by asking how the same models presented in
Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
expression. To make contact with experiment, all of these models must make a
commitment to some numerical values for the key parameters found in each such
model. Therefore in Section~\ref{section_04_bayesian_inference.tex} we explore
the use of Bayesian inference to establish these parameters and to rigorously
answer the question of how to discriminate between the different models.

%\mmnote{Key ideas, no particular order, that I haven't written down before:
%\begin{itemize}
%\item Our goal is to build phenomenological models of input-output functions of
%genetic circuits. More precisely, we want phenomenology in the sense of
%coarse-graining away atomistic detail, but still retaining biophysical meaning,
%e.g., we don't want to coarse-grain as far as Hill functions. Why not? We are
%motivated by studies like~\cite{Razo-Mejia2020}, for which Hill functions are
%insufficient. Such work requires a quantitative theory of how biophysical
%changes at the molecular level propagate to input-output functions at the
%genetic circuit level. We want concepts, not mere facts. In particular a key
%question is: at this level of coarse graining, what microscopic details do I
%need to explicitly model, and how do we figure that out? For example, do I need
%to worry about all or even any of the steps that individual RNAPs go through
%each time they make a transcript? Turning the question around, can we see any
%imprint of those processes in the available data? If the answer is no, then
%those processes are irrelevant for our purposes. Forward modeling and inverse
%(statistical inferential) modeling can complement each other beautifully here.
%\end{itemize}}
